---
title: "Homework 1"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

<!-- knitr global options -->

```{r, include = FALSE}
knitr::opts_chunk$set(comment=NA, fig.width=4.5, fig.height=3.5, fig.align = 'center')
```

# Introduction

This problem set will replicate a well-known paper on racial bias in the
labor market: "Are Emily and Greg More Employable Than Lakisha and
Jamal? A Field Experiment on Labor Market Discrimination" by Marianne
Bertrand and Sendhil Mullainathan. The paper, which I'll refer to as BM
for short, appears in Volume 94, Issue \#4 of the *American Economic
Review* (AER). Before beginning this lab, visit the website of the
*American Economic Review* and search for the paper. Once you've found
it, download a pdf of the paper. Alternatively, you can also download
this paper by searching for it in the UPenn library.

# Exercise \#1

Read the introduction and conclusion of BM, and then write a
one-paragraph summary addressing the following points:

1.  What question does the paper try to answer?
2.  What data and methodology are used to address the question?
3.  What are the key findings?

# Solution to Exercise \#1

1.  The paper tries to answer if there is still a differential racial
    treatment by race in the U.S. labor market.

2.  Data is being collected from job applicants resume outside Boston
    and Chicago newspapers. There are few variables such as :

    1)  college degree (dummy),

    2)  years of experience,

    3)  volunteering experience (dummy),

    4)  military experience (dummy),

    5)  email address (dummy),

    6)  employment holes (dummy),

    7)  work in school (dummy),

    8)  honors (dummy),

    9)  computer skills (dummy),

    10) special skills (dummy),

    11) fraction high school dropouts in applicant's zip code,

    12) fraction college or more in applicant's zip code,

    13) fraction Whites in applicant's zip code,

    14) fraction African Americans in applicant's zip code

    15) long (median per-capita income) in applicant's zip code

3.  <div>

    -   A white applicant have a higher chance for the callback for
        every 10 ads job vacancy they applied for, compared to african
        american if they want to get a callback they have to apply for
        15 ads job vacancy.

    -   The paper found that a higher quality resumes of white-named
        applicant, they will receive more callback rates compared to a
        low quality resumes of white-named applicant. The same effects
        applied as well for the african-american named applicant, but
        with fewer callback rates compared to the white-named one. Three
        times larger improved callback rates for the white-named
        applicant compared to african-american named applicant.
        In-short, the white-named applicant receive more benefits in
        terms of improved resumes quality.

    -   The neighborhood and address location doesn't contribute for the
        callback rates. There is no significant callback rates for
        african-american people that lives in the whiter, more educated
        zip code compared to not live in that neighborhood. This one
        applies for the white-named applicant as well.

    -   There is a little systematic relationship between the
        requirements of the job and the racial gap in callback and the
        variables do not show any statistically significant.

    </div>

# Importing the Dataset

The dataset is in Canvas as `lakisha_aer.dta`. The extension `dta`
implies that this is a STATA file. STATA is a commercial statistics
software with a proprietary file format. Fortunately, some enterprising
open-source programmers have written software that can decode `.dta`
files and convert them into other formats. We'll use the function
`read_dta()` to convert `lakisha_aer.dta` into a tibble that we can
manipulate with `dplyr`. This function is in the package `haven`, which
also contains functions for converting data from SPSS, STATA, and SAS
formats. Make sure to install `haven` before proceeding.

Note that you'll have to specify the directory where you've saved the
data. Below, I use `setwd()` to specify the directory. The path is for a
Mac. If you're using Windows, it's a little trickier to specify the
right file path: you may need to Google this.

Load the `.dta` file as a tibble into R using `read_dta()`. Call your
tibble `ResumeNames`.

```{r}
library(tidyverse)
library(haven)
setwd("/Users/damarega/Documents/Econ224/") 
ResumeNames = read_dta('lakisha_aer.dta')
```

Each row in `ResumeNames` corresponds to a single fictitious job
applicant. For instance, `sex` and `race` refer to the gender and
ethnicity of the (fictitious) individual in the resume.

# Exercise \#2

1.  Display the tibble `ResumeNames`. How many rows and columns does it
    have?
2.  Display only the columns `sex`, `race` and `firstname` of
    `ResumeNames`.
3.  Add two new columns to `ResumeNames`: `female` should take the value
    `TRUE` if `sex` is female, and `black` should take value `TRUE` if
    `race` is "b".

# Solution to Exercise \#2

```{r}
#number 1 
tibble(ResumeNames)
#It has 65 columns and 4870 rows
```

```{r}
#number 2
ResumeNames %>%
  select(sex, race, firstname)

```

```{r}
ResumeNames %>%
  mutate(female = case_when(
    sex == "f" ~ "TRUE"
  )) %>%
  mutate(black = case_when(
    race == "b" ~ "TRUE"
  )) %>%
  select(female, black)
```

# Checking for Balance

The dataset is an example of a Randomized Control Trial (RCT). The
authors randomly sent out fictitious "White sounding" and "Black
sounding" resumes to employers in Boston and Chicago.

The first thing to check is whether the information mentioned in the
resumes: `education`, `ofjobs`, `sex`, `computerskills` and `yearsexp`
is similar across both sets of resumes. This is known as *Checking for
Balance*, and I will refer to this in the context of the current
experiment as checking for balance across race. Checking for balance is
important to ensure everything else is held constant apart from the
ethnicity of the applicant. That way any difference in call backs can be
attributed entirely to discrimination.

The variable `computerskills` takes on the value `1` if a given resume
says that the applicant has computer skills. The variables `education`
and `yearsexp` indicate level of education and years experience, while
`ofjobs` indicates the number of previous jobs listed on the resume.

You should answer all the relevant coding questions using `dplyr`.

# Exercise \#3

1.  Read parts A-D of section II in BM and answer the following:

    -   How did the experimenters create their bank of resumes for the
        experiment?
    -   The experimenters classified the resumes into two groups. What
        were they and how did they make the classification?
    -   How did the experimenters generate identities for their
        fictitious job applicants?

2.  Is sex balanced across race? Hint: what happens if you apply the
    function `sum` to a vector of `TRUE` and `FALSE` values?

3.  Are computer skills balanced across race? Hint: the summary
    statistic you'll want to use is the *proportion* of individuals in
    each group with computer skills.

4.  Compute the mean and standard deviation of `yearsexp` by race.
    Comment on your findings.

5.  Are `education` and `ofjobs` balanced across race?

6.  Is `education` balanced across `sex`? Does it matter whether it is
    balanced?

# Solution to Exercise \#3

*Write your code and solutions here.*

**Number 1**

A. The experimenter use the resumes of actual job searchers but they
alter them to create distinct resumes. They done this by gather the
resumes needed on two job search websites. To differ the datasets of
resumes that they create with the real resumes job applicants, they
eliminate the resumes sent out in the Boston and Chicago area. They also
restrict four occupational categories : sales, administrative report,
clerical services, and customer services. Not only that, they restrict
resumes that has been posted for more than 6 months prior to the start
of the experiment.

B. They classified the resumes into two groups : high and low quality
resumes. To classify them, they use criteria such as labor market
experience, career profile, existence of gaps in employment, and skills
listed. And to make the gap further, they add to each
high-qualityresumea subsetof the following features: summer or
while-at-school employment experience, volunteeringexperience,
extracomputerskills, certificationdegrees, foreign language skills,
honors, or some military experience.

C. The researcher generated their fictitious job applicants by using
some data from the real job applicants. The research itself aiming to
unveil if there is a job discrimination across race for those who want
to apply for a job. Thus, the impression of a name plays an important
part. To give an impression of name, the researcher must differentiate
between the white-sounding names and black-sounding names. In order to
do that, the researcher use name from birth certificates of all babies
that were born in Massachusetts between 1974 and 1979. Thus, they can
create a new identities but based on a real name that doesn't distort
the impression for the employer. 

How they differentiate a white-sounding name from a black one? The
researcher conduct a survey in various public areas in Chicago. Each of
the respondent was given a name and asked to assess features of the
person, one of which being race. In general, the names led respondents
to readily attribute the expected race for the person but there were a
few exceptions and these names were disregarded. Thus, by using these
name, it will give the impression of the real white-sounding and
black-sounding names for the employer.

```{r}
#solution for number 2
ResumeNames %>%
  mutate(gender = case_when(
    sex == "f" ~ 1,
    sex == "m" ~ 1
  )) %>%
  group_by(race,sex) %>%
  summarize(total_sex_by_races = sum(gender))
  
#From this result, we can see that the white_male is only one third of the white_female population, so we can conclude that for the white race, its not balanced. The same applies in the black race case.
```

```{r}
#solutions for number 3
ResumeNames %>%
  group_by(race) %>%
  summarize(total_comp_skills_by_race = sum(computerskills))
#The computer skills between race are different. White race have fewer computer skills compared to black race in this datasets.
```

```{r}
#solutions for number 4
ResumeNames %>%
  group_by(race) %>%
  summarize(mean_yearsexp = mean(yearsexp), sd_yearsexp = sd(yearsexp))
#From the table below, we can conclude that the mean years experience among the race seems pretty balanced despite that white race have more average years of experience compared to black race. In standard deviation, black race have lower sd compared to white race.
```

```{r}
#solutions for number 5
ResumeNames %>%
  group_by(race) %>%
  summarize(edu_by_race = sum(education), totaljobslisted_by_race = sum(ofjobs))
#From the table below, we can conclude that education by race and total jobs listed by race is pretty balanced among race.
```

# Callbacks by Race and Sex

The outcome of interest in `ResumeNames` is `call` which takes on the
value `1` if the corresponding resume elicits an email or telephone
callback for an interview. Check your results in this section against
Table 1 of the paper.

# Exercise \#4

1.  Calculate the average callback rate for all resumes in
    `ResumeNames`.
2.  Calculate the average callback rates separately for resumes with
    "white-sounding" and "black-sounding" names. What do your results
    suggest?
3.  Repeat part 2, but calculate the average rates for each combination
    of race and sex. What do your results suggest?

# Solution to Exercise \#4

*Write your code and solutions here.*

```{r}
#number 1 solutions
mean(ResumeNames$call)
```

```{r}
#number 2 solutions
ResumeNames %>%
  mutate(names_by_races = case_when(
    race == "w" ~ "white-sounding names",
    race == "b" ~ "black-sounding names"
  )) %>%
  group_by(names_by_races) %>%
  summarize(callback_rates = mean(call))
#From this calculation, we know that the average callback rate for white-sounding name was higher compared to the average callback rate for black-sounding name. This proved that white-sounding name has more privilege compared to black-sounding name in terms of applying for a job cavancy. For white-sounding names, they will get 1 callback for every 11 jobs they applied, but in case of black-sounding names, they only get 1 callback for every 15 jobs they applied.

  
```

```{r}
#number 3 solutions (belom kelar)
ResumeNames %>%
  mutate(names_by_races = case_when(
    race == "w" ~ "white-sounding names",
    race == "b" ~ "black-sounding names"
  )) %>%
  group_by(names_by_races, sex) %>%
  summarize(callback_rates = mean(call))
#From the table below, we can see that female job applicants have higher callback rates compared to the male job applicant in either black or white race.

```

# Comparing Returns to Quality

Bertrand and Mullainathan write: "for most of the employment ads we
respond to, we send four different resumes: two higher-quality and
two-lower quality ones." The column `h` takes on the value `1` if a
resume as classified *a priori* as "high quality" and `0` if it was
classified as "low quality." The columns `col`, `military`, `email`,
`volunteer` are indicators for: college degree, has an email address,
has done volunteer work, and served in the military.

# Exercise \#5

1.  Compare the average value of `col`, `military`, `email`, and
    `volunteer` across "high quality" and "low quality" resumes. Discuss
    your findings.
2.  Calculate average callback rates for black versus white-sounding
    names *separately* for "high-quality" and "low-quality" resumes.
    Discuss your findings

# Solution to Exercise \#5

*Write your code and solutions here.*

```{r}
#Number 1
ResumeNames %>%
  mutate(quality = case_when(
    h == 1 ~ "high",
    l == 1 ~ "low",
  )) %>%
  group_by(quality) %>%
  summarize(mean_col=mean(col), mean_mil=mean(military), mean_email=mean(email), mean_volunteer=mean(volunteer))
#From the table below, we can see that either high quality and low quality applicants have college degree in average. We can see the different when it comes to military, email, and volunteering aspects. The high quality applicants have way more military and volunteering experiences and also they have emails compared to the low quality applicants.
```

```{r}
ResumeNames %>%
  mutate(names_by_race = case_when(
    race == "w" ~ "white-sounding names",
    race == "b" ~ "black-sounding names",
  )) %>%
  mutate(quality = case_when(
    h == 1 ~ "high",
    l == 1 ~ "low",
  )) %>%
group_by(names_by_race, quality) %>%
summarize(callbackrates = mean(call))
#From the table below, we can conclude that there is a lower effect for black-sounding names despite they have a high quality or low quality resumes despite white-sounding names.
```
